Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-03 numerical.tex
almost done with version 2

\section{Numerical solutions}\label{numerical}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the thin-disk, low-frequency and nearly-Keplerian
approximations used in the analytical discussion above, and  
use the full expressions of $\rho$, $\Omega$ and
$\kappa$ given in \S\ref{eqm}. We also restore $\hat{g}_c=1$ to
account for the background radial disk structure.

@@

We solve the linear problem by expanding the
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. At the vertical boundaries we impose a free
surface, 
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
where $\bm{\xi}$ is the Lagrangian displacement. This boundary
condition is convenient for numerical implementation since it conforms
to a standard eigenvalue problem. The above discretization procedure
converts the linear system of differential equations to a set of 
algebraic equations, for which we use matrix routines in the LAPACK
package to solve.

@@

\subsection{VSI in a locally isothermal, neutrally stratified disk}\label{vertiso_pertiso} 
% disk model is a nearly-vertically
% isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
% $\epsilon=0.1$. 
We first calculate the VSI in a nearly vertically isothermal, neutrally
stratified disk ($N_z^2\equiv0$) by setting
$\gamma=\Gamma=1.011$ and $\beta=10^{-3}$. Other disk parameters are 
$(p,q,\epsilon)=(-1.5,-1,0.1)$ and the vertical domain is $\zmax =
5\epsilon r$. This setup is similar to that adopted in
\cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).  This
allows us to test our numerical code.

@@

% This means
% that our analytic results should only be applied to perturbations with
% $\khat\gg 1$, i.e. small radial-lengthscales.

@@

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in \S\ref{iso_poly}. For completeness we show the
corresponding eigenfunctions in vertical velocity in
Fig. \ref{lowfreq_eigenfuncvz}, which show that the fundamental mode
consists of zero nodes in $\delta v_z$.

@@

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental
VSI mode, with one node in $W$, has the smallest $|\sigma|$. The
eigenvalue calculated from Eq. \ref{sig2_iso} with $L=1$ and the 
numerically-obtained value are  
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2606)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{sig2_iso})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surprisingly good given the number of approximations
used to obtain Eq. \ref{sig2_iso}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrals with $\rho$ as the   weight function, which rapidly
decays away from the midplane.

@@

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{Growth rate of the fundamental VSI as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The vertical line
     is the critical thermal timescale $\beta_\mathrm{crit}$ for the
     fundamental mode according to Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the eigenfunction $W$ for
$\beta=0.01$ and $\beta=0.1$. The latter value is close to the
critical value beyond which the fundamental VSI is stabilized
(\S\ref{iso_vsi_beta_crit}). Increasing $\beta$ 
restricts  the region in which $W\sim z$ closer to the midplane, with
oscillatory behavior emerging at the boundaries, where buoyancy first
becomes important. Finite thermal relaxation increases $|W|$ near the
vertical boundaries relative to that about the midplane, but
perturbations near the disk surface is unlikely important in practice
because the disk atmosphere contains little mass.

@@

\subsection{Critical thermal relaxation timescale}
% may need to explain why choose k=10
% bcrit independent of k for k>>1 
% for some parameter regimes (e.g. large q, large epsilon) 
% behavior at larger kx and finite bcool not understood. 
We test the dependence of the critical thermal relaxation timescale 
$\beta_\mathrm{crit}$ for the fundamental VSI on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the previous case study as a
reference point and vary $q\in[-0.2,-1.2]$, 
$\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
separately. Here, we use a wavenumber $\khat=10$, for which we
find growth rates reach zero for the range of disk parameters
considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely.

@@

\subsection{Vertically non-isothermal disks} 
We briefly consider vertically non-isothermal disks with 
$\Gamma=1.4$ and $(p,q,\epsilon)=(0,-1,0.05)$ as simulated in
\cite{nelson13}. In this case we set $\zmax\simeq2.2\epsilon r$, which
is close to the zero-density surface.

@@

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the fundamental VSI mode. This is
similar to that observed for nearly vertically isothermal disks
(Fig. \ref{bcrit_compare1}). For $\gamma=1.7,\,2.0,\,2.5$, growth
rates reach zero at  $\beta\simeq0.17,\,0.083,\,0.045$, respectively.
Interestingly, these values are equal to
$\epsilon|q|/(\gamma-\Gamma)$. This suggests that the critical thermal
relaxation timescale for vertically non-isothermal disks can also be
estimated by Eq. \ref{iso_vsi_cond} but with $\gamma-1$ replaced by
$\gamma-\Gamma$.

@@

% %To select the appropriate
% %eigenvalues we vary the parameters slowly away from a reference solution.   
% % The numerically-determined eigenfrequencies are qualitatively
% % consistent with the discussion in \S\ref{adia_compare_gam1}.
% There is good agreement between Fig. \ref{adia_growth_num} and
% Fig. \ref{adia_growth} for $\khat=k_xH_\mathrm{iso}\gtrsim1$. There is a
% mismatch for smaller wavenumbers, especially in the growth
% rates. Actual growth rates decrease slower as $\khat\to0$
% than that in Fig. \ref{adia_growth}. Disagreement at small
% $\khat$ is not unexpected because the low-frequency
% approximation used to obtain Fig. \ref{adia_growth} becomes invalid since
% $|\omega|\sim\Omega_k$ for small $\khat$.

>>>>

\section{Numerical solutions}\label{numerical}
In this section we numerically solve the linear stability problem
corresponding to Eq. \ref{lin_mass}---\ref{lin_energy}. 
We relax the thin-disk, low-frequency and nearly-Keplerian
approximations used in the analytical discussion above, and  
use the full expressions of $\rho$, $\Omega$ and
$\kappa$ given in \S\ref{eqm}. We also set $\hat{g}_c=1$ to
account for the background radial disk structure.

@@

We solve the linear problem by expanding the
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. At the vertical boundaries we impose a free
surface, 
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
where $\bm{\xi}$ is the Lagrangian displacement, the meridional
components of which are given by $\xi_{x,z} = \ii\delta
v_{x,z}/\sigma$. Calculations with a rigid boundary ($\delta v_z=0$)
yield similar results. 
% This boundary
% condition is convenient for numerical implementation since it conforms
% to a standard eigenvalue problem.

The above discretization procedure
converts the linear system of differential equations to a set of 
algebraic equations, for which we use matrix routines in the LAPACK
package to solve.

@@

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
% disk model is a nearly-vertically
% isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
% $\epsilon=0.1$. 
We first calculate the VSI in a neutrally
stratified disk ($N_z^2\equiv0$) with rapid thermal relaxation by
setting $\gamma=\Gamma=1.011$ and $\beta=10^{-3}$. Other disk
parameters are $(p,q,\epsilon)=(-1.5,-1,0.1)$ and the vertical domain
is $\zmax = 5H$. The disk is nearly vertically isothermal with
$c_s(\zmax)\simeq 0.94 c_s(0)$. Fig. \ref{omega_z} shows the
equilibrium rotation profile away from the disk mid-plane.

This setup is similar to that adopted
in \cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).  This
allows us to test our numerical code.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/omega2} 
  \caption{Equilibrium rotation profile $\Omega(z)$,
    normalized by its mid-plane value, for a disk with $\Gamma=1.011$
    and $(p,q,\epsilon)=(-1.5,-1,0.1)$. 
    \label{omega_z} 
  }
\end{figure}

@@

{\bf ASK. Growth rates are expected to diminish for both $\khat\to 0$
  and $\khat\to \infty$. This is because for $\khat\ll 1$ the radial
  wavelenth is large, but such disturbances are 
  stabilized by by epicyclic motions ($|\omega|\sim
  \kappa$). Disturbances with $\khat\gg 1$ are radially localized, but  
  there is no energy change when fluid elements are perturbed at fixed
  cylindrical radius whilst conserving its angular momentum (since we
  consider axisymmetric perturbations).  
}

@@

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in \S\ref{iso_poly}. For completeness we show the
corresponding eigenfunctions in vertical velocity in
Fig. \ref{lowfreq_eigenfuncvz}. The fundamental mode
consists of zero nodes in $\delta v_z$.

@@

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental has the
smallest $|\sigma|$, for which the eigenvalue calculated from
Eq. \ref{sig2_iso} with $L=1$ and the numerically-obtained value are   
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2606)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{sig2_iso})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surprisingly good given the number of approximations
used to obtain Eq. \ref{sig2_iso}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrals with $\rho$ as the   weight function, which rapidly
decays away from the midplane.

@@

 \begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare2} 
   \caption{Growth rate of the fundamental VSI as
     a function of the thermal relaxation timescale $\beta$. The disk
     parameters are $\Gamma=1.011$, $\gamma=1.4$, and 
     $(p,q,\epsilon)=(-1.5,-1,0.05)$. The vertical line
     is the critical thermal timescale $\beta_\mathrm{crit}$  obtained
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

@@

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the vertical velocity
perturbation $\delta v_z$ for
$\beta=0.01$ and $\beta=0.1$. The latter value is close to the
critical value beyond which the fundamental VSI is stabilized
(\S\ref{iso_vsi_beta_crit}). Increasing $\beta$ 
restricts  the region in which $\delta v_z\sim $ constant closer to the midplane, with
oscillatory behavior emerging at the boundaries, where buoyancy first
becomes important. Finite thermal relaxation increases $|\delta v_z|$ near the
vertical boundaries relative to that about the midplane. However, 
perturbations near the disk surface may not be energetically important
in practice because the disk atmosphere contains little mass.

@@

\subsection{Critical thermal relaxation timescale}\label{bcrit_num_test}
% may need to explain why choose k=10
% bcrit independent of k for k>>1 
% for some parameter regimes (e.g. large q, large epsilon) 
% behavior at larger kx and finite bcool not understood. 
We test the dependence of the critical thermal relaxation timescale 
$\beta_\mathrm{crit}$ for the fundamental VSI on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the previous case study as a
reference point and vary $q\in[-0.2,-1.2]$, 
$\gamma\in[1.2,2.0]$, and $\epsilon\in[0.02,0.1]$
separately. Here, we use a wavenumber $\khat=10$, for which we
find growth rates reach zero for the range of disk parameters
considered, so that a numerical $\beta_\mathrm{crit}$ can be defined precisely.

@@

\subsection{Vertically non-isothermal disks} 
We briefly consider vertically non-isothermal disks with 
$\Gamma=1.4$ and $(p,q,\epsilon)=(0,-1,0.05)$ as simulated in
\cite{nelson13}. In this case we set $\zmax\simeq2.2H$, which
is close to the zero-density surface.

@@

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the fundamental mode. This is
similar to that observed for nearly vertically isothermal disks
(Fig. \ref{bcrit_compare1}). For $\gamma=1.7,\,2.0,\,2.5$, growth
rates reach zero at  $\beta\simeq0.17,\,0.083,\,0.045$, respectively.
Interestingly, these values are equal to
$\epsilon|q|/(\gamma-\Gamma)$. This suggests that the critical thermal
relaxation timescale for vertically non-isothermal disks can also be
estimated by Eq. \ref{iso_vsi_cond} but with $\gamma-1$ replaced by
$\gamma-\Gamma$.

@@

% %To select the appropriate
% %eigenvalues we vary the parameters slowly away from a reference solution.   
% % The numerically-determined eigenfrequencies are qualitatively
% % consistent with the discussion in \S\ref{adia_compare_gam1}.
% There is good agreement between Fig. \ref{adia_growth_num} and
% Fig. \ref{adia_growth} for $\khat=k_xH\gtrsim1$. There is a
% mismatch for smaller wavenumbers, especially in the growth
% rates. Actual growth rates decrease slower as $\khat\to0$
% than that in Fig. \ref{adia_growth}. Disagreement at small
% $\khat$ is not unexpected because the low-frequency
% approximation used to obtain Fig. \ref{adia_growth} becomes invalid since
% $|\omega|\sim\Omega_k$ for small $\khat$.

### 2015-03-03 paper.tex
almost done with version 2

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the
  disk  midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in a stably stratified disk. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include an energy
  equation with a thermal relaxation term that restores the disk
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN)
  with an estimate of the physical thermal timescales in the MMSN
  based on dust opacity. We find the MMSN is unstable to the VSI at
  tens of AU with a characteristic growth times of tens of local
  orbital periods. This suggests the VSI is dynamically important in
  the outer parts of protoplanetary disks.  
\end{abstract}

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depends on the height from the
  disk  midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in a stably stratified disk. %which is usually assumed 
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we generalize the linear, axisymmetric VSI in radially local but
  vertically global models of accretion disks to include an energy
  equation with a thermal relaxation term that restores the disk
  temperature $T$ over a characteristic timescale
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response  of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from
  zero can dramatically reduce VSI growth rates. % , since this generally
  % introduces a vertical entropy gradient that stabilizes vertical
  % motions as needed for the VSI.
  Our  linear calculations are consistent with previous hydrodynamic
  simulations. For vertically isothermal disks, we quantify the
  requirement for `fast' thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. We apply our linear calculations 
  to a disk model of the Minimum Mass Solar Nebulae (MMSN) with an 
  estimate for the physical thermal timescales 
  in the MMSN based on dust opacity. We find the MMSN is unstable to
  the VSI at tens of AU with a characteristic growth times of $\sim
  30$ local orbital periods.  
  % This suggests the VSI is dynamically important in
  % the outer parts of protoplanetary disks.  
\end{abstract}

### 2015-03-03 setup.tex
updating main part of paper
almost done with version 2

We also consider disk equilibria in which the pressure $P$ and density
$\rho$ are related by   
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a  constant parameter  and the
function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives 
locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
sound-speed. %For $\Gamma=\gamma$ the disk is adiabatically stratified
% in the vertical 
It shall be convenient to define the function $c_s(r,z)$ 
\begin{align}
  c_s^2\equiv % \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
  \frac{\Gamma P}{\rho}, 
\end{align}
which is related to the isothermal sound-speed $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$. We take $c_{s0}(r)$ to be another power-law:  
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index, and
define   
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k$ is the Keplerian azimuthal velocity and
$H$ is a characteristic scale-height.  Thus, $\epsilon$
is also a characteristic disk aspect-ratio.  In practice, we consider
thin disks with $\epsilon \ll 1$.

For discussion purposes it is useful to define the vertical buoyancy
frequency $N_z$ 
\begin{align}
  N_z^2 \equiv - c_p^{-1}g_z \frac{\p S}{\p z} = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
where $c_p$ is the heat capacity at constant pressure, $g_z\equiv
\rho^{-1}\p_zP $ and $S\equiv c_p^{-1}\ln{P^{1/\gamma}/\rho}$ is the
specific entropy. The second equality applies to the disk equilibria
described above. 
We are interested in
neutrally or sub-adiabatically stratified 
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

>>>>

We consider disk equilibria in which the pressure $P$ and density
$\rho$ are related by   
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a  constant parameter  and the
function $c_{s0}(r)$ is prescribed below. Choosing $\Gamma=1$ gives 
locally vertically isothermal disk and $c_{s0}(r)$ is the mid-plane
sound-speed. %For $\Gamma=\gamma$ the disk is adiabatically stratified
% in the vertical 
It shall be convenient to define the function $c_s(r,z)$ 
\begin{align}
  c_s^2\equiv % \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} =
  \frac{\Gamma P}{\rho}, 
\end{align}
which is related to the isothermal sound-speed $c_\mathrm{iso} =
c_s/\sqrt{\Gamma}$. We take $c_{s0}(r)$ to be another power-law:  
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index, and
define   
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k$ is the Keplerian azimuthal velocity and
$H$ is a characteristic scale-height.  Thus, $\epsilon$
is also a characteristic disk aspect-ratio.  In practice, we consider
thin disks with $\epsilon \ll 1$.

For discussion purposes it is useful to define the vertical buoyancy
frequency $N_z$ via
\begin{align}
  N_z^2 \equiv - c_p^{-1}g_z \frac{\p S}{\p z} = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
where $c_p$ is the heat capacity at constant pressure, $g_z\equiv
\rho^{-1}\p_zP $ and $S\equiv c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy. The second equality applies to the disk equilibria
described above. 
We are interested in
neutrally or sub-adiabatically stratified 
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.
